`phyr`: An R package for model-based phylogenetic analyses

### 2019-06-24 ms.Rmd
update [skip ci]

comm = phyr::comm_a
set.seed(1234)
comm = comm[sample(nrow(comm), 5), sample(ncol(comm), 5)]
comm$site = row.names(comm)
dat = tidyr::gather(comm, key = "sp", value = "freq", -site) %>% 
  left_join(phyr::envi, by = "site") %>% 
  left_join(phyr::traits, by = "sp")
dat$pa = as.numeric(dat$freq > 0)
dat$freq2 = 20 - dat$freq
pglmm_example = phyr::communityPGLMM(freq ~ 1 + shade + (1 | sp__) + (1 | site) + (1 | sp__@site), 
                     data = dat, tree = phyr::phylotree)
```

@@

1. Phylogenetic relationships among species are both a challenge and an opportunity for community ecology and comparative biology. Model-based phylogenetic analysis has advantages ... Currently, R functions to conduct these analyses were distributed across different packages and were mostly written in R, which can be very slow with large datasets.
2. To help remedy this situation, we created `phyr`, an R package that collects and updates functions to conduct model-based phylogenetic analyses. By collecting these functions into one package and rewrite core functions in c++, this package can facilitate the use of model-based phylogenetic analyses.
3. This paper presents the main functions of the package and provides an example using simulated data.
4. The `phyr` package provides a unified software environment with improved performance to conduct model-based phylogenetic analyses.

@@

(ref:maunFuncCaption) Main functions in `phyr`.

@@

## Species' characteristics

@@

### Speed comparisons

@@

# Limitations

`communityPGLMM` requires columns `site` and `sp`.

# Conclusion

>>>>

comm = phyr::comm_a
comm$site = row.names(comm)
# add site environmental data and species traits
dat = tidyr::gather(comm, key = "sp", value = "freq", -site) %>% 
  left_join(phyr::envi, by = "site") %>% 
  left_join(phyr::traits, by = "sp")
dat$pa = as.numeric(dat$freq > 0)
dat$freq2 = 20 - dat$freq
pglmm_example = phyr::communityPGLMM(freq ~ 1 + shade + (1 | sp__) + (1 | site) + (1 | sp__@site), 
                     data = dat, cov_ranef = list(sp = phyr::phylotree))
```

@@

1. Phylogenetic relationships among species are both a challenge and an opportunity for community ecology and comparative biology. Model-based phylogenetic analysis has advantages ... Currently, R functions to conduct these analyses were distributed across different packages and were mostly written in R, which can be very slow with large datasets.
2. To help remedy this situation, we created `phyr`, an R package that collects and updates functions to conduct model-based phylogenetic analyses. By collecting these functions into one package and rewrite core functions in c++, this package can facilitate the use of model-based phylogenetic analyses.
3. This paper presents the main functions of the package and provides an example using simulated data.
4. The `phyr` package provides a unified software environment with improved performance to conduct model-based phylogenetic analyses in Ecology.

@@

(ref:maunFuncCaption) List of main functions in `phyr` package.

@@

## Species' traits

@@

@@

```{r}
tree = phyr::phylotree
phyr::psv(comm_b, tree) # comm_b is a site by species matrix
phyr::psd(comm_b, tree)
x = phyr::pcd(comm_b, tree, verbose = FALSE)
```

## Species' traits

```{r}
pglmm_example
```

# Benchmark

```{r, echo=FALSE}
# data prep for pez::communityPGLMM, not necessary for phyr::communityPGLMM
dat = arrange(dat, site, sp)
nspp = n_distinct(dat$sp)
nsite = n_distinct(dat$site)

dat$site = as.factor(dat$site)
dat$sp = as.factor(dat$sp)

tree = ape::drop.tip(phylotree, setdiff(phylotree$tip.label, unique(dat$sp)))
Vphy <- ape::vcv(tree)
Vphy <- Vphy/max(Vphy)
Vphy <- Vphy/exp(determinant(Vphy)$modulus[1]/nspp)
Vphy = Vphy[levels(dat$sp), levels(dat$sp)]

# prepare random effects
re.site <- list(1, site = dat$site, covar = diag(nsite))
re.sp <- list(1, sp = dat$sp, covar = diag(nspp))
re.sp.phy <- list(1, sp = dat$sp, covar = Vphy)
# sp is nested within site
re.nested.phy <- list(1, sp = dat$sp, covar = Vphy, site = dat$site)
re.nested.rep <- list(1, sp = dat$sp, covar = solve(Vphy), site = dat$site) # equal to sp__@site
# can be named 
re = list(re.sp = re.sp, re.sp.phy = re.sp.phy, re.nested.phy = re.nested.phy, re.site = re.site)
```

```{r eval=T}
microbenchmark::microbenchmark(phyr::pcd(comm = comm_a, tree = phylotree, reps = 1000, verbose = F),
                               picante::pcd(comm = comm_a, tree = phylotree, reps = 1000),
                               times = 30)
rt = ape::rtree(n = 1000)
microbenchmark::microbenchmark(phyr::vcv2(rt, corr = T),
                               ape::vcv(rt, corr = T))

microbenchmark::microbenchmark(phyr::psv(comm_a, phylotree, compute.var = T),
                               picante::psv(comm_a, phylotree, compute.var = T))

microbenchmark::microbenchmark(
  phyr::communityPGLMM(pa ~ 1 + shade + (1|sp__) + (1|site) + (1|sp__@site), dat, 
                       family = "binomial", cov_ranef = list(sp = phylotree), REML = F, cpp = T, 
                       optimizer = "bobyqa"),
    phyr::communityPGLMM(pa ~ 1 + shade + (1|sp__) + (1|site) + (1|sp__@site), dat, 
                       family = "binomial", cov_ranef = list(sp = phylotree), REML = F, cpp = T, 
                       optimizer = "Nelder-Mead"),
  phyr::communityPGLMM(pa ~ 1 + shade + (1|sp__) + (1|site) + (1|sp__@site), dat, 
                       family = "binomial", cov_ranef = list(sp = phylotree), REML = F, cpp = F, 
                       optimizer = "Nelder-Mead"),
  pez::communityPGLMM(pa ~ 1 + shade, data = dat, family = "binomial", sp = dat$sp, 
                      site = dat$site, random.effects = re, REML = F),
  times = 10
)
```

# Implications

# Resources

This package is available at Github.

# Future directions

In the future, we intend to ...
